---
title: "PART 2: Signature profile visualization and analysis"
author: ["Shixiang Wang, Ziyu Tao, Tao Wu, Xue-Song Liu (Corresponding author)"]
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    toc_depth: 3
bibliography: ref.bib
link-citations: yes
---


```{r setup-02, include=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print = "75")
opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  prompt = FALSE,
  tidy = "styler",
  comment = NA,
  dpi = 300,
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
opts_knit$set(width = 75)
Sys.setenv("LANGUAGE" = "EN")
```

In this part, we would like to visualize and analyze all tumor samples as a whole.

## Signature profile visualization

```{r}
library(tidyverse)
library(sigminer)

pcawg_sigs <- readRDS(file = "../data/pcawg_cn_sigs_CN176_signature.rds")
```

Show signature profile with relative contribution value for each component.

```{r, fig.height=13, fig.width=14}
show_sig_profile(
  pcawg_sigs,
  mode = "copynumber",
  method = "X",
  style = "cosmic"
)
```

Show signature profile with absolute contribution value (estimated segment count) for each component.

```{r, fig.height=13, fig.width=14}
show_sig_profile(
  pcawg_sigs,
  mode = "copynumber",
  method = "X",
  style = "cosmic",
  normalize = "raw"
)
```

Show signature activity profile.

```{r fig.width=14, fig.height=7}
show_sig_exposure(
  pcawg_sigs,
  style = "cosmic",
  rm_space = TRUE,
  palette = sigminer:::letter_colors
)
```

## Signature analysis

### Signature similarity

Here we explore the similarity between copy number signatures identified in PCAWG data. High cosine similarity will cause some issues in signature assignment and activity attribution [@mauraPracticalGuideMutational2019].

```{r}
sim <- get_sig_similarity(pcawg_sigs, pcawg_sigs)
```

```{r, fig.width=5.5, fig.height=5}
pheatmap::pheatmap(sim$similarity, cluster_cols = F, cluster_rows = F, display_numbers = TRUE)
```

Most of signature-pairs have similarity around `0.1`.

### Signature activity distribution

Transform data to long format.

```{r}
df_abs <- get_sig_exposure(pcawg_sigs) %>% 
  tidyr::pivot_longer(cols = starts_with("CNS"), names_to = "sig", values_to = "activity") %>% 
  dplyr::mutate(activity = log10(activity + 1),
                sig = factor(sig, levels = paste0("CNS", 1:14)))

df_rel <- get_sig_exposure(pcawg_sigs, type = "relative", rel_threshold = 0) %>% 
  tidyr::pivot_longer(cols = starts_with("CNS"), names_to = "sig", values_to = "activity") %>% 
  dplyr::mutate(
    sig = factor(sig, levels = paste0("CNS", 1:14))
  )
```

Plot with function in `sigminer`.

```{r}
show_group_distribution(
  df_abs,
  gvar = "sig",
  dvar = "activity",
  order_by_fun = FALSE,
  g_angle = 90,
  ylab = "log10(activity+1)"
)
```

```{r}
show_group_distribution(
  df_rel,
  gvar = "sig",
  dvar = "activity",
  order_by_fun = FALSE,
  g_angle = 90,
  ylab = "relative activity"
)
```

## Copy number profile and signature activity

### Copy number profile for representative samples

Here we will draw copy number profile of 2 representative samples for each signature.

```{r}
get_sample_from_sig <- function(dt, sig) {
  res = head(dt[order(dt[[sig]], dt[[paste0("ABS_", sig)]], decreasing = TRUE)], 2L)
  res
}
```

Read `CopyNumber` object for PCAWG data.

```{r}
pcawg_cn_obj <- readRDS("../data/pcawg_cn_obj.rds")
samp_summary <- pcawg_cn_obj@summary.per.sample
rel_activity <- get_sig_exposure(pcawg_sigs, type = "relative", rel_threshold = 0)
abs_activity <- get_sig_exposure(pcawg_sigs, type = "absolute")

rel_activity <- rel_activity[, lapply(.SD, function(x) {if (is.numeric(x)) round(x, 2) else x})]
# all(rownames(abs_activity) == rownames(rel_activity))
colnames(abs_activity) <- c("sample", paste0("ABS_CNS", 1:14))
act <- merge(
  rel_activity, abs_activity,
  by = "sample"
)
```


```{r, fig.width=12, fig.height=7}
dir.create("../output/enrich_samples/", showWarnings = FALSE)
for (i in paste0("CNS", 1:14)) {
  cat(paste0("Most enriched in ", i, "\n"))
  s = get_sample_from_sig(act, i)
  print(s)
  plist = show_cn_profile(pcawg_cn_obj, samples = s$sample,
                          show_title = TRUE,
                          return_plotlist = TRUE)
  plist = purrr::map2(plist, s$sample, function(p, x) {
    s = samp_summary[sample == x]
    text = paste0(
      "n_of_seg:", s$n_of_seg, "\n",
      "n_of_amp:", s$n_of_amp, "\n",
      "n_of_del:", s$n_of_del, "\n",
      "rel:", act[sample == x][[i]], "\n",
      "abs:", act[sample == x][[paste0("ABS_", i)]], "\n"
    )
    p <- p + annotate("text", x = Inf, y = Inf, hjust = 1, vjust = 1,
                      label = text, color = "gray50")
    p
  })
  p = cowplot::plot_grid(plotlist = plist, nrow = 2)
  print(p)
  ggplot2::ggsave(file.path("../output/enrich_samples/", paste0(i, ".pdf")),
                    plot = p, width = 12, height = 6)
}
```

Plot sample profile for presentation in manuscript.

```{r}
sel_samps <- c("SP117655", "SP123894", "SP105708", "SP116525", "SP117309",
               "SP77443", "SP114903", "SP116539", "SP118048", "SP112887",
               "SP135354", "SP37636", "SP43664", "SP96163")
# table(pcawg_cn_obj@data[sample == "SP116539"]$chromosome)
# table(pcawg_cn_obj@data[sample == "SP112887"]$chromosome)
plist = list()
for (s in sel_samps) {
  if (s == sel_samps[4]) {
    p = show_cn_profile(pcawg_cn_obj, samples = s, chrs = c("chr12"),
                        show_title = TRUE,
                        return_plotlist = TRUE)
  } else if (s == sel_samps[8]) {
    p = show_cn_profile(pcawg_cn_obj, samples = s, chrs = c("chr1", "chr2", "chr4", "chr5", "chr6"),
                        show_title = TRUE,
                        return_plotlist = TRUE)
  } else if (s == sel_samps[10]) {
    p = show_cn_profile(pcawg_cn_obj, samples = s, chrs = paste0("chr", c(1:10, 14, 16, 18, 19, 21)),
                        show_title = TRUE,
                        return_plotlist = TRUE)
  } else if (s == sel_samps[13]) {
    p = show_cn_profile(pcawg_cn_obj, samples = s, chrs = c("chr7"),
                        show_title = TRUE,
                        return_plotlist = TRUE)
  } else {
    p = show_cn_profile(pcawg_cn_obj, samples = s,
                        show_title = TRUE,
                        return_plotlist = TRUE)
  }
  plist[[s]] <- p[[1]] + ggplot2::labs(x = NULL)
  
}

plist = purrr::pmap(list(plist, sel_samps, paste0("CNS", 1:14)), function(p, x, y) {
  s = samp_summary[sample == x]
  text = paste0(
    "n_of_seg:", s$n_of_seg, "\n",
    "n_of_amp:", s$n_of_amp, "\n",
    "n_of_del:", s$n_of_del, "\n",
    "rel:", act[sample == x][[y]], "\n",
    "abs:", act[sample == x][[paste0("ABS_", y)]], "\n"
  )
  p <- p + annotate("text", x = Inf, y = Inf, hjust = 1, vjust = 1,
                    label = text, color = "gray50")
  p
})
p = cowplot::plot_grid(plotlist = plist, ncol = 1)
ggplot2::ggsave(file.path("../output/enrich_samples/", "selected_samples.pdf"),
                plot = p, width = 12, height = 42)
```



### Ideograph for copy number profile reconstruction

Select one/two samples and draw copy number profile -> catalog profile -> signature relative activity.

```{r}
ss <- c("SP102561", "SP123964")
```

```{r}
p = show_cn_profile(pcawg_cn_obj, samples = ss,
                        show_title = TRUE, ncol = 1)
ggplot2::ggsave("../output/example_cn_profile.pdf",
                plot = p, width = 12, height = 6)
```

```{r}
tally_X <- readRDS("../data/pcawg_cn_tally_X.rds")
```

```{r}
p = show_catalogue(tally_X, mode = "copynumber", method = "X", style = "cosmic", samples = ss, by_context = TRUE, font_scale = 0.7)
ggplot2::ggsave("../output/example_sig_profile.pdf",
                plot = p, width = 15, height = 4)
```

```{r}
p <- show_sig_exposure(sig_exposure(pcawg_sigs)[, ss], hide_samps = FALSE) + ggplot2::ylab("Activity") + ggpubr::rotate_x_text(0, hjust = 0.5)
ggplot2::ggsave("../output/example_sig_activity.pdf",
                plot = p, width = 4, height = 5)
```




```{r, include=FALSE}
# Clear environment
rm(list = ls())
```

