---
title: "PART 6: Benchmark analysis"
author: ["Ziyu Tao, Shixiang Wang, Chenxu Wu, Huimin Li, Tao Wu, Xiangyu Zhao, Wei Ning, Guangshuai Wang, Xue-Song Liu (Corresponding author)"]
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    toc_depth: 3
bibliography: ref.bib
link-citations: yes
---


```{r setup-06, include=FALSE}
library(knitr)
library(rmdformats)
library(sigminer)
library(ggplot2)
library(tidyverse)
## Global options
options(max.print = "75")
opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  prompt = FALSE,
  tidy = "styler",
  comment = NA,
  dpi = 300,
  message = FALSE,
  warning = FALSE,
  echo = TRUE
)
opts_knit$set(width = 75)
Sys.setenv("LANGUAGE" = "EN")
```

To evaluate the performance of our method across different platform and copy number calling tools. We carried out benchmark analysis in prostate cancer.

## Compare the copy number calling algorithms

For WGS, we select copy number from ABSOLUTE and ACEseq. For WES, we select copy number from ABSOLUTE and ASCAT.  

```{r message=FALSE}
tally_SNP_ABSOLUTE <- readRDS("../data/benchmark/tally_SNP_ABSOLUTE_matrix.rds")
tally_SNP_ASCAT2 <- readRDS("../data/benchmark/tally_SNP_ASCAT2_matrix.rds")


# Compare different CNV algorithms in SNP
SNP_sample <- intersect(rownames(tally_SNP_ABSOLUTE), rownames(tally_SNP_ASCAT2))
tally_SNP_ABSOLUTE <- subset(tally_SNP_ABSOLUTE, rownames(tally_SNP_ABSOLUTE) %in% SNP_sample)
tally_SNP_ASCAT2 <- subset(tally_SNP_ASCAT2, rownames(tally_SNP_ASCAT2) %in% SNP_sample)
sim_profile.SNP <- sapply(SNP_sample, function(i){
    dt1.1 <- tally_SNP_ASCAT2[i,]
    dt2.1 <- tally_SNP_ABSOLUTE[i,]
    sigminer:::cosine(dt1.1, dt2.1)
}) %>% 
  subset(. > 0.4)
median(sim_profile.SNP)
```

```{r}
hist(sim_profile.SNP, breaks = 50, xlab = "",
     main = "Copy number profile similarity from two methods(SNP)",xlim = range(0,1))
abline(v = 0.8500122, col = "red",lty=2)
```

We compared the signature between tools in SNP. 

```{r message=FALSE}
SNP_ABSOLUTE_SP_sigs <- readRDS("../data/benchmark/SNP_ABSOLUTE_SP_highmatch.rds")
SNP_ASCAT2_SP_sigs <- readRDS("../data/benchmark/SNP_ASCAT2_SP_highmatch.rds")
colnames(SNP_ABSOLUTE_SP_sigs$solution_list$S4$Signature.norm) <- paste0("SNP_ABSOLUTE_sigs",seq(1:4))
colnames(SNP_ASCAT2_SP_sigs$solution_list$S4$Signature.norm) <- paste0("SNP_ASCAT_sigs",seq(1:4))
sim <- get_sig_similarity(SNP_ABSOLUTE_SP_sigs$solution_list$S4, SNP_ASCAT2_SP_sigs$solution_list$S4)
```

```{r fig.width=6, fig.height=4}
p <- pheatmap::pheatmap(sim$similarity, cluster_cols = F, cluster_rows = F, display_numbers = TRUE)
```

```{r}
tally_WGS_ABSOLUTE <- readRDS("../data/benchmark/tally_WGS_ABSOLUTE_matrix.rds")
tally_WGS_aceseq <- readRDS("../data/benchmark/tally_WGS_aceseq_matrix.rds")

# Compare different CNV algorithms in WGS
WGS_sample <- intersect(rownames(tally_WGS_ABSOLUTE), rownames(tally_WGS_aceseq))
tally_WGS_ABSOLUTE <- subset(
  tally_WGS_ABSOLUTE, 
  rownames(tally_WGS_ABSOLUTE) %in% WGS_sample
  )
tally_WGS_aceseq <- subset(tally_WGS_aceseq, rownames(tally_WGS_aceseq) %in% WGS_sample)
sim_profile.WGS <- sapply(WGS_sample, function(i){
  dt1.1 <- tally_WGS_ABSOLUTE[i,]
  dt2.1 <- tally_WGS_aceseq[i,]
  sigminer:::cosine(dt1.1, dt2.1)
}) %>% 
  subset(. > 0.4)
hist(sim_profile.WGS, breaks = 50, xlab = "",
     main = "Copy number profile similarity from two methods(WGS)",xlim = range(0,1))
abline(v = 0.9725061, col = "red",lty=2)
```

We compared the signature between tools in WGS.  

```{r message=FALSE}
WGS_ABSOLUTE_SP_sigs <- readRDS("../data/benchmark/WGS_ABSOLUTE_SP_highmatch.rds")
WGS_aceseq_SP_sigs <- readRDS("../data/benchmark/wgs_aceseq_sig.rds")
colnames(WGS_ABSOLUTE_SP_sigs$solution_list$S4$Signature.norm) <- paste0("WGS_ABSOLUTE_sigs",seq(1:4))
colnames(WGS_aceseq_SP_sigs$Signature.norm) <- paste0("WGS_ACEseq_sigs",seq(1:4))
sim <- get_sig_similarity(WGS_ABSOLUTE_SP_sigs$solution_list$S4, WGS_aceseq_SP_sigs)
```

```{r fig.width=6, fig.height=4}
p <- pheatmap::pheatmap(sim$similarity, cluster_cols = F, cluster_rows = F, display_numbers = TRUE)
```

## Compare the data platforms

CNA signatures have been extracted independently from WGS, WES and SNP array derived prostate cancer CNA profiles. We compare the 4 copy number signatures extracted by Sigprofiler platform. 

### SNP vs WGS

```{r message=FALSE}
SNP_sigs_source <- readRDS("../data/benchmark/SNP_ABSOLUTE_SP_source.rds")
WGS_sigs_source <- readRDS("../data/benchmark/WGS_ABSOLUTE_SP_source.rds")
WES_sigs_source <- readRDS("../data/benchmark/WES_FACETS_SP_source.rds")

colnames(SNP_sigs_source$solution_list$S4$Signature.norm) <- paste0("SNP_sigs",seq(1:4))
colnames(WGS_sigs_source$solution_list$S4$Signature.norm) <- paste0("WGS_sigs",seq(1:4))
colnames(WES_sigs_source$solution_list$S4$Signature.norm) <- paste0("WES_sigs",seq(1:4))
sim <- get_sig_similarity(SNP_sigs_source$solution_list$S4, WGS_sigs_source$solution_list$S4)
```

```{r fig.width=6, fig.height=4}
p <- pheatmap::pheatmap(sim$similarity, cluster_cols = F, cluster_rows = F, display_numbers = TRUE)
```

### SNP vs WES

```{r fig.width=6, fig.height=4}
sim <- get_sig_similarity(SNP_sigs_source$solution_list$S4, WES_sigs_source$solution_list$S4)
p <- pheatmap::pheatmap(
  sim$similarity, 
  cluster_cols = F, 
  cluster_rows = F, 
  display_numbers = TRUE,
  legend_breaks=c(0.2,0.4,0.6,0.8)
  )
```
### WES vs WGS

```{r fig.width=6, fig.height=4}
sim <- get_sig_similarity(WES_sigs_source$solution_list$S4, WGS_sigs_source$solution_list$S4)
p <- pheatmap::pheatmap(sim$similarity, cluster_cols = F, cluster_rows = F, display_numbers = TRUE)
```
